---
title: "One-sided"
output: 
  bookdown::html_document2:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{one-sided}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`RTSA` is intended as a tool for sequential meta-analysis using Trial Sequential Analysis (TSA) in the statistical software R. To ensure the accuracy and quality of the code in `RTSA`, we will be comparing the output of RTSA to other packages. Unfortunately, no package in R offer sequential stopping boundaries for meta-analysis. This is one of the reasons for creating the `RTSA` package. Therefore, we will sometimes be comparing results to the original TSA software coded in Java and sometimes to packages on CRAN not intended specifically for sequential meta-analysis.

The methods for calculating the sequential stopping boundaries in TSA are almost identical to the methods for calculating sequential stopping boundaries in clinical trials. Hence we can compare the `RTSA` code with the existing packages in R for calculating stopping boundaries. We will specifically be comparing our stopping boundaries to the `gsDesign` package in R although `gsDesign` is created for sequential methods in clinical trials and `RTSA` is created for sequential methods in meta-analysis.

Besides doing sequential meta-analysis, `RTSA` can also perform traditional meta-analysis. As the package is not created mainly for this purpose, we recommend using the `metafor` package in R for traditional meta-analysis as this package has an extensive selection of methods and very detailed documentation. The `metafor` package function also worked as an inspiration for some of the functions implemented in `RTSA` We will in this vignette be comparing to the `metafor` package when conducting traditional meta-analyses.

# Setup

We will be using the `RTSA`, `metafor` and `gsDesign` packages in this vignette.

```{r message=FALSE}
library(RTSA)
library(gsDesign)
library(metafor)
```

```{r, include=FALSE}
library(dplyr)
library(kableExtra)
```

We will use a data example which is included in the `RTSA` package called `perioOxy`.

```{r}
data("perioOxy")
head(perioOxy)
```

# Traditional meta-analysis

A traditional meta-analysis (non-sequential meta-analysis) can be computed in `RTSA`.

In `RTSA`, we will be using the functions `metaPrepare` and `synthesize` to create a traditional meta-analysis. The function metaPrepare takes binary summary data (events and number of participants) as input and returns a list of trial specific treatment effects, confidence intervals and more. See Table \@ref(tab:trameta) for a presentation of some of the results from the metaPrepare function fit.

```{r}
mp = RTSA:::metaPrepare(outcome = "RR", data = perioOxy, method = "IV", alpha = 0.05)
```

The escalc function in the metafor package does essentially the same as the metaPrepare function. To compute confidence intervals per trial, the summary function can be used. The results from the escalc function can be seen in Table \@ref(tab:trameta).

```{r}
me = escalc(measure="RR", n1i=nI, n2i=nC, ai=eI, ci=eC, data=perioOxy)
sum.me = summary(me)
```

```{r trameta, echo = FALSE}
out.mat = cbind(perioOxy$study, round(mp$te, 2), paste0("(", round(mp$lower,2), "; ", round(mp$upper,2),")"), round(exp(me$yi),2), paste0("(", round(exp(sum.me$ci.lb),2), "; ", round(exp(sum.me$ci.ub),2),")"))
colnames(out.mat) = c("Trial", "Effect estimate ", "95% CI", "Effect estimate ", "95% CI")
knitr::kable(out.mat, caption = "Results from the metaPrepare and escalc function") %>% kable_paper() %>% add_header_above(c("", "RTSA::metaPrepare" = 2, "metafor::escalc and summary" = 2))
```

The saved object from metaPrepare can then be used to compute a meta-analysis made by the function synthesize in RTSA. The function returns a list containing information about a fixed-effect meta-analysis, a random-effects meta-analysis and more. We will specifically be fitting a fixed-effect model (always fitted) and a random-effects model (fitted per default) using the DerSimonian-Laird estimator for the heterogeneity. We will be comparing the output of the synthesize function with the metafor package function rma. See Table \@ref(tab:trametaout).

```{r}
sm = RTSA:::synthesize(mp, sign = NULL, fixedStudy = FALSE,hksj = FALSE, tau.ci.method = "BJ") # RTSA function
out.mtFE = rma(yi, vi, data = me, method = "FE") # metafor function
out.mtRE = rma(yi, vi, data = me, method = "DL") # metafor function
```

```{r trametaout, echo = FALSE}
out.mat = rbind(c(round(sm$peF[c(1,2,3)],2),round(sm$peF[5],4),
                  c(round(exp(c(out.mtFE$beta, out.mtFE$ci.lb, out.mtFE$ci.ub)),2),
                    round(out.mtFE$pval,4))),
                c(round(sm$peR[c(1,2,3)],2),round(sm$peR[5],4),
                  round(exp(c(out.mtRE$beta, out.mtRE$ci.lb, out.mtRE$ci.ub)),2),
                  round(out.mtRE$pval,4)))
out.mat[,c(2,6)] = c(paste0("(", out.mat[,2], "; ", out.mat[,3],")"),
                    paste0("(", out.mat[,6], "; ", out.mat[,7],")"))
out.mat = out.mat[,-c(3,7)]
colnames(out.mat) = c("Effect estimate ", "95% CI", "p-value", "Effect estimate ",
                      "95% CI", "p-value")
rownames(out.mat) = c("Fixed-effect", "Random-effects")
knitr::kable(out.mat, caption = "Results from the synthesize and rma function") %>% kable_paper() %>% add_header_above(c("", "RTSA::synthesize" = 3, "metafor::rma" = 3))
```

# Sequential meta-analysis

The RTSA package is mainly created for sequential meta-analysis. We will first focus on the boundary calculation before looking at the main RTSA function in our package.

```{r}
# Calculate the cumulative number of participants
count <- cumsum(perioOxy$nI+perioOxy$nC)

# Calculate the RIS 
RR = 0.9
p0 = sum(perioOxy$eC+perioOxy$eI)/sum(perioOxy$nC+perioOxy$nI)
pI = exp(log(p0)+log(RR))
pC = exp(log(p0)-log(RR))
RIS = RTSA:::nRandom(alpha = 0.05, beta = 0.2, pI = pI, pC = pC, diversity = 0.2)

# Set the timings of the studies relative to the RIS
timing <- c(count/RIS,1)

# create increment in the information
timingincr <- timing - c(0,timing[-length(timing)])
trials <- cbind(timing, timingincr, c(count,RIS))

# Start BoundaryCalculations programme ------------------------------------
# do not analyse after too small an increment in information
IFincrementThreshold <- 0.01
IFtotalThreshold <- 0.05

trials <- trials[trials[,2] > IFincrementThreshold,]

# calculate the boundaries
boundaries <- RTSA:::boundary(inf_frac = trials[,1],
                 side = 1, alpha = 0.05, zninf = -8, tol = 1e-9)
boundaries
```

We can compare our boundaries to the gsDesign package.

```{r}
gs = gsDesign(k = 7,
         n.I = trials[,3],
         test.type = 1,
         alpha = 0.05, sfu = sfLDOF, sfl = sfLDOF)
gs$upper$bound
```

```{r}
gs = gsDesign(k = 7,
         n.I = trials[,3],
         test.type = 1,
         alpha = 0.05, sfu = sfLDOF, sfl = sfLDOF)
gs$lower$bound
```

## Plot

```{r, fig.height = 8, fig.width = 8}
outRTSA <- RTSA(data = perioOxy, outcome = "RR", mc = 0.7, side = 1, alpha = 0.05)
plot(outRTSA)
```
