---
title: "Sequential boundaries in RTSA and other packages in R"
output: 
  bookdown::html_document2:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{other-software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

RTSA is intended as a tool for sequential meta-analysis using Trial Sequential Analysis (TSA) in statistical software R. To ensure the correctness and quality of the code in RTSA, we will be comparing the output of RTSA to other packages. Unfortunately, no package in R offer sequential stopping boundaries for meta-analysis. This is one of the reasons for creating the RTSA package. For this reason we will sometimes be comparing results to the original TSA software coded in java and sometimes to packages on CRAN not intended for sequential meta-analysis. 

The methods for calculating the sequential stopping boundaries in TSA are almost identical to the methods for calculating sequential stopping boundaries in clinical trials. Hence we can compare the RTSA code with the existing packages in R for calculating stopping boundaries. We will specifically be comparing our stopping boundaries to the gsDesign package in R although gsDesign is created for sequential methods in clinical trials and RTSA is created for sequential methods in meta-analysis. 

Besides doing sequential meta-analysis, RTSA can also perform traditional meta-analysis. As the package is not created mainly for this purpose, we recommend using the metafor package in R for traditional meta-analysis as this package has an extensive selection of methods and documentation. The metafor package function also worked as an inspiration for some of the functions implemented in RTSA. We will in this vignette be comparing to the metafor package when conducting traditional meta-analyses.

# Setup

We will be using the RTSA, metafor and gsDesign packages in this vignette. 

```{r message=FALSE}
library(RTSA)
library(gsDesign)
library(metafor)
```

```{r, include=FALSE}
library(dplyr)
library(kableExtra)
```


We will use a data example which is included in the RTSA package called perioOxy.

```{r}
data("perioOxy")
head(perioOxy)
```

# Traditional meta-analysis
A traditional meta-analysis (non-sequential meta-analysis) can be computed in RTSA. 

In RTSA, we will be using the functions metaPrepare and synthesize to create a traditional meta-analysis. The function metaPrepare takes binary summary data (events and number of participants) as input and returns a list of trial specific treatment effects, confidence intervals and more. See Table \@ref(tab:trameta) for a presentation of some of the results from the metaPrepare function fit. 

```{r}
mp = metaPrepare(outcome = "RR", eI = perioOxy$eI, nI = perioOxy$nI,
                 eC = perioOxy$eC, nC = perioOxy$nC, method = "IV")
```

The escalc function in the metafor package does essentially the same as the metaPrepare function. To compute confidence intervals per trial, the summary function can be used. The results from the escalc function can be seen in Table \@ref(tab:trameta).

```{r}
me = escalc(measure="RR", n1i=nI, n2i=nC, ai=eI, ci=eC, data=perioOxy)
sum.me = summary(me)
```

```{r trameta, echo = FALSE}
out.mat = cbind(perioOxy$trial, round(mp$te, 2), paste0("(", round(mp$lower,2), "; ", round(mp$upper,2),")"), round(exp(me$yi),2), paste0("(", round(exp(sum.me$ci.lb),2), "; ", round(exp(sum.me$ci.ub),2),")"))
colnames(out.mat) = c("Trial", "Effect estimate ", "95% CI", "Effect estimate ", "95% CI")
knitr::kable(out.mat, caption = "Results from the metaPrepare and escalc function") %>% kable_paper() %>% add_header_above(c("", "RTSA::metaPrepare" = 2, "metafor::escalc and summary" = 2))
```

The saved object from metaPrepare can then be used to compute a meta-analysis made by the function synthesize in RTSA. The function returns a list containing information about a fixed-effect meta-analysis, a random-effects meta-analysis and more. We will specifically be fitting a fixed-effect model (always fitted) and a random-effects model (fitted per default) using the DerSimonian-Laird estimator for the heterogeneity. We will be comparing the output of the synthesize function with the metafor package function rma. See Table \@ref(tab:trametaout). 

```{r}
sm = synthesize(mp) # RTSA function
out.mtFE = rma(yi, vi, data = me, method = "FE") # metafor function
out.mtRE = rma(yi, vi, data = me, method = "DL") # metafor function
```

```{r trametaout, echo = FALSE}
out.mat = rbind(c(round(sm$peF[c(1,2,3)],2),round(sm$peF[5],4),
                  c(round(exp(c(out.mtFE$beta, out.mtFE$ci.lb, out.mtFE$ci.ub)),2),
                    round(out.mtFE$pval,4))),
                c(round(sm$peR[c(1,2,3)],2),round(sm$peR[5],4),
                  round(exp(c(out.mtRE$beta, out.mtRE$ci.lb, out.mtRE$ci.ub)),2),
                  round(out.mtRE$pval,4)))
out.mat[,c(2,6)] = c(paste0("(", out.mat[,2], "; ", out.mat[,3],")"),
                    paste0("(", out.mat[,6], "; ", out.mat[,7],")"))
out.mat = out.mat[,-c(3,7)]
colnames(out.mat) = c("Effect estimate ", "95% CI", "p-value", "Effect estimate ",
                      "95% CI", "p-value")
rownames(out.mat) = c("Fixed-effect", "Random-effects")
knitr::kable(out.mat, caption = "Results from the synthesize and rma function") %>% kable_paper() %>% add_header_above(c("", "RTSA::synthesize" = 3, "metafor::rma" = 3))
```

# Sequential meta-analysis

The RTSA package is mainly created for sequential meta-analysis. We will first focus on the boundary calculation before looking at the main RTSA function in our package. 

```{r}
# Calculate the cumulative number of participants
count <- cumsum(perioOxy$nI+perioOxy$nC)

# Calculate the RIS 
RR = 0.9
p0 = sum(perioOxy$eC+perioOxy$eI)/sum(perioOxy$nC+perioOxy$nI)
pI = exp(log(p0)+log(RR))
pC = exp(log(p0)-log(RR))
RIS = nRandom(alpha = 0.05, beta = 0.2, pI = pI, pC = pC, diversity = 0.2)

# Set the timings of the studies relative to the RIS
timing <- c(count/RIS,1)

# create increment in the information
timingincr <- timing - c(0,timing[-length(timing)])
trials <- cbind(timing, timingincr, c(count,RIS))

# Start BoundaryCalculations programme ------------------------------------
# do not analyse after too small an increment in information
IFincrementThreshold <- 0.01
IFtotalThreshold <- 0.05

trials <- trials[trials[,2] > IFincrementThreshold,]

# calculate the boundaries
boundaries <- boundary(informationFractions = trials[,1],
                 side = 2, alpha = 0.05, zninf = -8, tol = 1e-24)
boundaries
```

We can compare our boundaries to the gsDesign package.

```{r}
gs = gsDesign(k = 7,
         n.I = trials[,3],
         test.type = 2,
         alpha = 0.025, sfu = sfLDOF, sfl = sfLDOF)
gs$upper$bound
gs$lower$bound
```

While the boundaries are not completely identical, they are very similar. It is also worth noting that the methods for calculating the boundaries are different. The method in gsDesign are largely based on the methods described in Jennison, Christopher and Turnbull, Bruce W., 2000, "Group Sequential Methods with Applications to Clinical Trials" ISBN: 0-8493-0316-8. See especially chapter 19 for a description of the numerical methods. The method for the RTSA boundary function is based on the java implementation different from the method in gsDesign. See Section "The history of the boundary calculation in RTSA" for more information.

To get an output with more informatin and less steps, we can use the RTSA function which provides different kinds of information. The information could be used for e.g. plotting as is done in the original TSA software. The function e.g. can set a number of analysis times (anaTimes) if we want to do a meta-analysis after the first 4, 5, 7 and 8 studies. 

```{r}
RTSA(timing = timing, synth = mp, anaTimes = c(4,5,7,8),
     side = 2, alpha = 0.05)
```
# Plotting in RTSA

The RTSA function can be used as an input for plotting the traditional TSA plot. We might consider making a plot function for RTSA objects. 

```{r}
RTSAout = RTSA(timing = timing, synth = mp, anaTimes = c(4,5,7,8),
     side = 2, alpha = 0.05)
plot(RTSAout)
```

# Extra examples just for the heck of it (TODO)

We show some more examples on other data sets. The next example shows that when the heterogeneity is estimated to 0, then the synout function will only print fixed-effect model results. 

```{r}
data("coronary")
mp = metaPrepare(outcome = "OR", eI = coronary$eI, nI = coronary$nI,
                 eC = coronary$eC, nC = coronary$nC, method = "MH")
synout = synthesize(mp, hakn = TRUE)
synout
```


# The history of the boundary calculation in RTSA

Originally the implementation of the boundary calculation method started with following chapter 19 in Jennison, Christopher and Turnbull, Bruce W., 2000, "Group Sequential Methods with Applications to Clinical Trials" ISBN: 0-8493-0316-8. This implementation was then tested against the gsDesign package, where the results were identical. For now, we are sticking with the TSA implementation to calculate boundaries, but we might do a larger investigation on which implementation is preferable. 
