---
title: "RTSA-manual"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Rtsa manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(RTSA)
```

# Introduction to Trial Sequential Analysis

Trial Sequential Analysis (TSA) is a method for conducting sequential meta-analysis.

The purpose of this manual is to function as a working manual for both statisticians and non-statisticians. It contains information about most of the key statistical methods behind TSA. For tutorials on how to use RTSA, we recommend reading the vignettes to this package. 

This manual will start with a short introduction to when a sequential meta-analysis should be considered before moving to main section of this manual about the statistical background. The statistical background section will touch upon hypothesis testing, type I and type II errors and more. 

This is a work in progress. Please refer to the original TSA manual at: 
https://ctu.dk/wp-content/uploads/2021/03/2017-10-10-TSA-Manual-ENG_ER.pdf

